---
title: "Project title"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

# Data 1

## Introduction and data

-   **Identify the source of the data.**

    -   The data is sourced from the National Center for Health Statistics (NCHS)

-   **State when and how it was originally collected (by the original data curator, not necessarily how you found the data).**

    -   Data are updated weekly. The provisional counts for COVID-19 deaths are based on a current flow of mortality data in the National Vital Statistics System. National provisional counts include deaths occurring within the 50 states and the District of Columbia that have been received and coded as of the date specified

-   **Write a brief description of the observations.**

    -   Each Observation is a COVID-19 deaths by age group, sex, week and state

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-1
covidData <- read_csv("data/Covid19Dataset.csv")
glimpse(covidData)
```

# Data 2

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-2
Athletes <- read_csv("data/Athletes.csv")
glimpse(Athletes)
```

# Data 3

## Introduction and data

-   Identify the source of the data.

    -   **California Safe Cosmetics Program in the California Department of Public Health**

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

    -   **The California Safe Cosmetics Act requires manufacturers, packers, and distributors of cosmetic products in California to report ingredients linked to health risks. Companies with over \$1 million in annual cosmetic sales and products sold in California since January 1, 2007, must comply. The data includes product details, chemical information, and reporting dates submitted to the California Safe Cosmetics Program.**

-   Write a brief description of the observations.

    -   

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-3
cosemetics <- read_csv("data/cscpopendata.csv")
glimpse(cosemetics)
```
