---
title: "Project title"
subtitle: "Report"
format: html
editor: visual
execute:
  echo: true
---

#### Introduction and data

The global outbreak of COVID-19 has created a need for comprehensive data to understand the impact that the virus held. This research seeks to uncover the paterns of Covid-19 related mortality in the United States in correlation with demographic factors. The data in this data set is collected and updated every week and covers information within all 50 states of the USA and the District of Columbia. The counts the number of COVID-19 deaths are based on data stored in the National Vital Statistic System. These observations are then added every week. Each Observation of the dataset represents the count of Covid-19 death in a given week, state, specific age group, and sex. So for every single week, there will be an observation for every combination of every subgroup (state, age, group, and sex). The data is sourced from the CDC's National Center for Health Statistics (NCHS, 2023). Our research question is: What is the relationship between age groups and COVID-19 mortality rates in the United States over time, and how have these rates evolved from the start of the pandemic until June 28, 2023? This research question allows us to see how Covid-19 mortality rates have changed over time among different age groups which helps us better understand the dynamic of the virus and how the pandemic potentially evolve. The research topic will be to observe trends in COVID-19 deaths given different geographic groups and how these trends evolve over time. A feasible hypothesis can be that those in older age groups will have higher mortality rate compared to the rest of the demographic. The types of variables in my research question include both Categorical and quantitative. Age in this context would be categorical because we are grouping age together and looking at specific age groups. The number of deaths in each observation is Quantitative. Sex and date are all categorical.

Given the nature of health data, it is essential to acknowledge and address potential ethical considerations. The data used in this dataset is annonymized and aggregate, which adheres to the highest standard of privacy. This allows the overarching goal of valuable significant insights while respecting the confidentiality of individuals involved in the dataset.

#### Methodology

The methodology section should include visualizations and summary statistics **relevant to your research question**. You are required to provide writing that explains the trends / patterns the visualization(s) displays as it relates to your research question. This writing can be treated as a "rough draft" for your final, but should also provide enough detail to **state and justify** the choice of statistical method(s) used to answer your research question.

**Research Question:** "What is the relationship between age groups and COVID-19 mortality rates in the United States over time, and how have these rates evolved from the start of the pandemic until June 28, 2023? What about the relationship between sex and COVID-19 mortality rate?"

```{r}
covidData <- read_csv("data/Covid19Dataset.csv")
```

```{r}
#｜label: take out all ages, keep only all sex
covidData_Ages <- covidData |>
  filter(!str_detect(`Age Group`,"All Ages"),str_detect(Sex, "All Sex"),ignore_case = TRUE)
covidData_Ages
```

Because the dataset contains all possible combinations of sex and age groups given a specific end week, we decided to drop all observations with "All Ages," which includes all ages in each observation, which would not allow us to visualize the relationship between COVID-19 deaths and specific age groups. Furthermore, we filtered the dataset to only include the Sex of "All Sex" because it already includes both the female and male data(Female data + male data = All Set Data). If we included the separate female and male observations, then the data would overlap. After dropping these rows in the "Ages" columns, we were able to graph our box plot visualization of the COVID-19 Deaths and each age group.

```{r}
covidData_Ages |>
ggplot(
    aes(
      y = `Age Group`, x = `COVID-19 Deaths`, fill = `Age Group`
    )
  ) +
  geom_boxplot() +
  labs(title = "Relationship between Age and COVID-19 Deaths",
       subtitle = "Deaths Per Week From 01/04/2020 to 06/24/2023")+
   theme(legend.position = "none")
```

From our visualization, we can conclude that that the median of COVID-19 Deaths of those 85 years and over is the highest. This trend stays mostly consistent throughout: as the age group increases, the median number of COVID-19 Deaths increase.

```{r}
#| label: take_out_all sex and keep only all ages
covidData_sex <- covidData |>
  filter(!str_detect(Sex,"All Sex"),str_detect(`Age Group`,"All Ages"),ignore_case=TRUE)

covidData_sex |>
  ggplot(aes(x = Sex, y = `COVID-19 Deaths`, fill = Sex)) +
    geom_bar(stat = "identity", width = 0.7) +
    labs(title = "Total COVID-19 Deaths by Sex",
         subtitle = "From 01/04/2020 to 06/24/2023",
         x = "Sex",
         y = "COVID-19 Deaths (Thousands)") +
    scale_y_continuous(labels = scales::comma_format(scale = 1e-3)) +
    theme(legend.position = "none")
```

For the same reason as above, we filtered the dataset to drop all sex and only keep rows that are all ages so there isn't repetition. We used a barplot to visualize the relationship between total deaths of male compared to females.


```{r}
#| label: linear regression equation for both models
modelsex <-linear_reg() |>
  set_engine("lm") |>
  fit(`COVID-19 Deaths` ~ Sex , data = covidData_sex)
glance(modelsex)$r.squared

modelage <- linear_reg() |>
  set_engine("lm") |>
  fit(`COVID-19 Deaths` ~ `Age Group` , data = covidData_Ages)
glance(modelage)$r.squared

```
#### Results

Provide results from your analysis that answer your research question. Reminder, the goal is not calculate every possible statistic and perform every possible procedure for all variables. You should demonstrate that you are proficient at asking meaningful questions and answering them using data, that you are skilled in interpreting and presenting results, and that you can accomplish these tasks using R. More is not better.

You are encouraged to provide writing that interprets your results and answers your research question. This includes writing a **scope of inference**. The writing is optional for the draft, but will be a requirement on the final.

#### Workflow + Formatting

To earn full credit, please be mindful of the following:

Have an updated `about` qmd for your project.

Responded to / Closed all *issues* from your proposal from chosen data set.

Your website must be able to be rendered by me and your lab leader prior to submission.

**EVERYONE** in your group must have at least 3 commits.

Pipes %\>%, \|\> and ggplot layers + should be followed by a new line.

All binary operators should be surrounded by space. For example x + y is appropriate. x+y is not.

All team members should contribute to the GitHub repository, with **regular meaningful commits.**
